---
title: "DataFest 2020"
subtitle: "Change in Flights Offered and Passengers Enplaned due to COVID-19 and Their Impact on the Airline Industry"
author: "Catherine Jennifer"
date: "5/5/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Step 1: Creating and cleaning the dataset


### 1A — COVID dataset

```{r}
covid_data <- read.csv("covid-us.csv")
head(covid_data)
tail(covid_data)
```


### 1B — Commercial and total flights datasets

```{r}
# Commercial flights
com_flights <- read.csv("commercial-flights.csv")
com_flights_old <- read.csv("commercial-flights-old.csv")
com_flights_old <- com_flights_old[1:9,]
com_flights <- rbind(com_flights_old, com_flights)
rm(com_flights_old)
colnames(com_flights) <- c("Date", "7_Day_Moving_Average", "Number_of_Flights")
head(com_flights)

# Total flights
total_flights <- read.csv("total-flights.csv")
total_flights_old <- read.csv("total-flights-old.csv")
total_flights_old <- total_flights_old[1:9,]
total_flights <- rbind(total_flights_old, total_flights)
rm(total_flights_old)
colnames(total_flights) <- c("Date", "7_Day_Moving_Average", "Number_of_Flights")
head(total_flights)
```


## Step 2: Exploratory data analysis


### 2A — Plotting `covid_data`

```{r}
# log of COVID cases
log_cases <- log(covid_data$cases)

# setting two semi-transparent colors
blue <- rgb(0, 0, 1, alpha = 0.5)
red <- rgb(1, 0, 0, alpha = 0.5)

barplot(log_cases, main = "Log Cases of COVID-19 in the United States",
        ylab = "log_cases",
        names.arg = covid_data$date,
        col = blue,
        border = NA,
        cex.names = 0.5,
        las = 2
)
```

```{r}
# log of COVID deaths
log_deaths <- log(covid_data$deaths)

# remove -Inf from converting to log scale
log_deaths[c(1:39)] <- 0

barplot(log_deaths, main = "Log Deaths of COVID-19 in the United States",
        ylab = "log_deaths",
        names.arg = covid_data$date,
        col = red,
        border = NA,
        cex.names = 0.5,
        las = 2
)

```

```{r}
# plot the two datasets together


barplot(log_cases, main = "Log Cases and Log Deaths of COVID-19 in the United States",
        col = blue,
        border = NA,
        space = 0,
        names.arg = covid_data$date,
        cex.names = 0.5,
        las = 2
)
barplot(log_deaths,
        col = red,
        border = NA,
        space = 0,
        las = 1,
        add = TRUE
)
legend("topleft", 
       legend = c("log_cases", "log_deaths"), 
       fill = c(blue, red),
       bty = "n")
```

**Explain why transforming case and death counts to the logarithmic scale makes sense.**


### 2B — Plotting `com_flights` and `total_flights`

```{r}
# setting two semi-transparent colors
cyan <- rgb(0, 0.5, 0.5, alpha = 0.5)
magenta <- rgb(0.5, 0, 0.5, alpha = 0.5)

# plotting com_flights
barplot(com_flights$Number_of_Flights, main = "Commercial Flights in the United States",
        names.arg = com_flights$Date,
        col = cyan,
        border = NA,
        cex.names = 0.5,
        las = 2
)
```

```{r}
# plotting total_flights
barplot(total_flights$Number_of_Flights, main = "Total Flights in the United States",
        names.arg = total_flights$Date,
        col = magenta,
        border = NA,
        cex.names = 0.5,
        las = 2
)
```

```{r}
par(mar = c(5,4,4,4) + 0.1)
# plotting com_flights and total_flights together
df_barplot <- barplot(total_flights$Number_of_Flights,
        main = "Flights in the United States",
        col = magenta,
        border = NA,
        space = 0,
        ylab = "Number of Flights",
        cex.lab = 0.75,
        names.arg = total_flights$Date,
        cex.names = 0.5,
        las = 2,
        cex.axis = 0.5,
        ylim = c(0,200000)
)
barplot(com_flights$Number_of_Flights,
        col = cyan,
        border = NA,
        space = 0,
        axes = FALSE,
        las = 1,
        add = TRUE
)

# only execute once
# master <- rbind(0.5, master)
par(new = TRUE)
plot(x = df_barplot, y = master$cases, col = "blue", lwd = 2, type = "l",
     axes = FALSE, xlab = NA, ylab = NA)
axis(side = 4, cex.axis = 0.5, las = 2)
mtext("Number of Cases/Deaths", side = 4, line = 3, cex = 0.75)
lines(x = df_barplot, y = master$deaths, col = "red", lwd = 2)

legend("top",
       legend = c("Total", "Commercial"),
       fill = c(magenta, cyan),
       bty = "n")
legend(65,1425000,
       legend = c("Cases","Deaths"),
       col = c("blue","red"),
       bty = "n",
       lwd = 2,
       lty = 1)
```


## Step 3: Comparing COVID and Flights


### 3A — Initial visualization

```{r}
barplot(total_flights$Number_of_Flights,
        main = "Flights in the United States",
        col = magenta,
        border = NA,
        space = 0,
        names.arg = total_flights$Date,
        cex.names = 0.5,
        las = 2,
        ylim = c(0,200000)
)
barplot(com_flights$Number_of_Flights,
        col = cyan,
        border = NA,
        space = 0,
        las = 1,
        add = TRUE
)
legend("topright", 
       legend = c("Total Flights", "Commercial Flights"), 
       fill = c(magenta, cyan),
       bty = "n")

plot(x = 1:113, y = log_cases, 
     main = "Log Cases and Log Deaths of COVID-19 in the United States",
     col = blue,
     type = "l",
     lty = 1,
     lwd = 2
)
# barplot(log_deaths,
#         col = red,
#         border = NA,
#         las = 1,
#         add = TRUE
# )
# legend("topleft", 
#        legend = c("log_cases", "log_deaths"), 
#        fill = c(blue, red),
#        bty = "n")
```


### 3B — Create master dataset

```{r}
master <- cbind(covid_data[16:113,], 
                com_flights$Number_of_Flights[1:98], 
                total_flights$Number_of_Flights[1:98])
colnames(master) <- c("date","cases","deaths","cflights","tflights")
row.names(master) <- NULL
head(master)
```


### 3C — Diagnose relationship between `cases` and `cflights`, i.e. if `cases` affect `cflights`

```{r}
# linear model, no transform
par(mfrow = c(1,2))
plot(cflights ~ cases, data = master, 
     main = "COVID-19 Cases vs. Commercial Flights",
     cex.main = 0.75)
cases_cflights_model <- lm(cflights ~ cases, data = master)
abline(cases_cflights_model, col = "blue")
plot(cases_cflights_model$residuals)
abline(h = 0, col = "red")

# linear, log transform
plot(cflights ~ log(cases), data = master, 
     main = "Log COVID-19 Cases vs. Commercial Flights",
     cex.main = 0.75)
cases_cflights_model2 <- lm(cflights ~ log(cases), data = master)
abline(cases_cflights_model2, col = "blue")
plot(cases_cflights_model2$residuals)
abline(h = 0, col = "red")

# linear, log-log transform
plot(log(cflights) ~ log(cases), data = master, 
     main = "Log COVID-19 Cases vs. Log Commercial Flights",
     cex.main = 0.75)
cases_cflights_model3 <- lm(log(cflights) ~ log(cases), data = master)
abline(cases_cflights_model3, col = "blue")
plot(cases_cflights_model3$residuals)
abline(h = 0, col = "red")

# diagnostic plots
par(mfrow = c(2,2))
plot(cases_cflights_model)
plot(cases_cflights_model2)
plot(cases_cflights_model3)

# summaries
options(digits = 8)
summary(cases_cflights_model)
summary(cases_cflights_model2)
summary(cases_cflights_model3)
```

From the plots above, we see that the best transformation out of the three is the log-log model: log `cases` and log `cflights`, which provides easier interpretability and greatly reduced residual standard error. However, the relationship between them are non-linear. Now, let's try fitting a quadratic and a cubic model to the log-log transformed variables.

### 3C — Fitting a quadratic and a cubic model

```{r}
# quadratic
quad_model <- lm(log(cflights) ~ log(cases) + I((log(cases))^2), data = master)
summary(quad_model)
par(mfrow = c(2,2))
plot(quad_model)

# cubic
cubic_model <- lm(log(cflights) ~ log(cases) + I((log(cases))^2) + 
                    I((log(cases))^3), data = master)
summary(cubic_model)
par(mfrow = c(2,2))
plot(cubic_model)
```

